# dbplyr {#dbplyr}

## Introduction

In this chapter, we will learn to query data from a database using dplyr.

We will use the following R packages:

- [DBI](https://dbi.r-dbi.org/)
- [RSQLite](https://rsqlite.r-dbi.org/)
- [dbplyr](http://dbplyr.tidyverse.org/)
- [dplyr](http://dplyr.tidyverse.org/)

```{r lite1a, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(dbplyr)
library(dplyr)
library(DBI)
library(RSQLite)
```

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/f5ee72cee9ab3256230cdedecd3ef25b).

## Connect to Database

Let us connect to an in memory SQLite database using `dbConnect()`.

```{r dbply1}
con <- dbConnect(RSQLite::SQLite(), ":memory:")
```

We will copy the `ecom` data to the database so that we can use it for running
dplyr statements. We read the same `ecom` data set used across chapters
(1000 rows, 8 columns including `country` and `purchase`) from the local
`data/web.csv` snapshot, falling back to the online copy if needed.

```{r dbply2, message=FALSE}
web_path <- if (file.exists("data/web.csv")) "data/web.csv" else "https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv"
ecom <- readr::read_csv(web_path, show_col_types = FALSE)
ecom <- dplyr::select(ecom, referrer, device, bouncers, n_visit, n_pages, duration, country, purchase)
dplyr::copy_to(con, ecom)
```

## Reference Data 

In order to use dplyr functions, we need to reference the table in the database using
`tbl()`.

```{r dbply3}
ecom2 <- dplyr::tbl(con, "ecom")
ecom2
```

## Query Data 

We will look at some simple examples. Let us start by selecting `referrer`, `device` and `duration` 
columns from `ecom2`.

```{r dbply4}
select(ecom2, referrer, device, duration)
```

We can filter data as well. Filter all the rows from `ecom2` where `duration` 
is greater than 300.

```{r dbply5}
filter(ecom2, duration > 300)
```

Time to do some grouping and summarizing. Let us compute the average time spent
on the site for different types of devices.

```{r dbply6}
ecom2 %>%
  group_by(device) %>%
  summarise(avg_duration = mean(duration))
```

## Show Query

If you want to view the SQL query generated in the above step, use `show_query()` or `explain()`.

```{r dbply7, collapse=TRUE}
avg_time <- 
  ecom2 %>%
  group_by(device) %>%
  summarise(avg_duration = mean(duration))

dplyr::show_query(avg_time)

dplyr::explain(avg_time)
```

## Collect Data

Now, some interesting facts. When working with databases, **dplyr** never pulls data into R unless you explicitly ask for it. In the previous example, dplyr will not do anything until you ask for the `avg_time` data. It generates the SQL and only pulls down a few rows when you try to print `avg_time`. 
So how do we pull all the data and store it for further analysis? `collect()` will pull all 
the data and store it in a tibble and you can use it for any further analysis.

```{r dbply9}
dplyr::collect(avg_time)
```
