# dbplyr {#dbplyr}

## Introduction

In this chapter, we will learn to query data from a database using dplyr.

We will use the following R packages:

- [DBI](http://readr.tidyverse.org/)
- [RSQLite](https://rstats-db.github.io/RSQLite/)
- [dbplyr](http://dbplyr.tidyverse.org/)
- [dplyr](http://dplyr.tidyverse.org/)

```{r lite1a, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(dbplyr)
library(dplyr)
library(DBI)
library(RSQLite)
```

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/f5ee72cee9ab3256230cdedecd3ef25b).

## Connect to Database

Let us connect to an in memory SQLite database using `dbConnect()`.

```{r dbply1}
con <- dbConnect(RSQLite::SQLite(), ":memory:")
```

We will copy the `mtcars` data to the database so that we can use it for running
dplyr statements.

```{r dbply2}
dplyr::copy_to(con, mtcars)
```

## Reference Data 

In order to use dplyr functions, we need to reference the table in the database using
`tbl()`.

```{r dbply3}
mtcars2 <- dplyr::tbl(con, "mtcars")
mtcars2
```

## Query Data 

We will look at some simple examples. Let us start by selecting `mpg`, `cyl` and `drat` 
columns from `mtcars2`.

```{r dbply4}
select(mtcars2, mpg, cyl, drat)
```

We can filter data as well. Filter all the rows from `mtcars2` where `mpg` is 
greater than 25.

```{r dbply5}
filter(mtcars2, mpg > 25)
```

Time to do some grouping and summarizing. Let us compute the average mileage for
different types of cylinders.

```{r dbply6}
mtcars2 %>%
  group_by(cyl) %>%
  summarise(mileage = mean(mpg))
```

## Show Query

If you want to view the SQL query generated in the above step, use `show_query()` or `explain()`.

```{r dbply7, collapse=TRUE}
mileages <- 
  mtcars2 %>%
  group_by(cyl) %>%
  summarise(mileage = mean(mpg))

dplyr::show_query(mileages)

dplyr::explain(mileages)
```

## Collect Data

Now, some interesting facts. When working with databases, **dplyr** never pulls data into R unless you explicitly ask for it. In the previous example, dplyr will not do anything until you ask for the mileages data. It generates the SQL and only pulls down a few rows when you try to print `mileages`. 
So how do we pull all the data and store it for further analysis? `collect()` will pull all 
the data and store it in a tibble and you can use it for any further analysis.

```{r dbply9}
dplyr::collect(mileages)
```
