# SQL Basics {#sqlbasics}

```{r setup, eval=TRUE, message=FALSE, echo=FALSE}
library(dplyr)
library(DBI)
library(RSQLite)
knitr::opts_chunk$set(R.options=list(max.print=50))
```

## Introduction

In this chapter, we will learn to:

- select 
    - single column
    - multiple columns
    - distinct values in a column
- limit the number of records returned
- handle NULL values
- and filter columns using the following operators
   - WHERE
   - AND, or & NOT
   - BETWEEN
   - IN
   - LIKE

We will use the following R packages:

- [DBI](http://readr.tidyverse.org/)
- [RSQLite](https://rstats-db.github.io/RSQLite/)
- [dbplyr](http://dbplyr.tidyverse.org/)

```{r lite1aa, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(dbplyr)
library(DBI)
library(RSQLite)
```

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/f44774d9c6e115c08d5e2e693380a00f).

## Set Up

In order to be able to use the `ecom` data set for our case study, we will read 
it using the **readr** package and then copy it to an in memory database using the 
`copy_to()` function from **dplyr**.

```{r setup2a, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, ecom)
```
## Select Columns

The SQL SELECT statement is used to fetch the data from a database table.

### Syntax

Below is the basic syntax of the SELECT statement.

`SELECT column1, column2, columnN FROM table_name;`

Here, column1, column2... are the fields of a table whose values you want to fetch. 
If you want to fetch all the fields, use the following syntax.

`SELECT * FROM table_name;`

### Select Single Column

Let us begin by selecting the `device` field from the `ecom` table.

```{sql, connection=con}
SELECT device FROM ecom
```

### Select Multiple Columns

Select the following fields from the `ecom` table:

- `referrer`
- `device`
- `purchase`

```{sql, connection=con}
SELECT referrer, device, purchase  FROM ecom
```

### Select All Columns

Select all the fields from the `ecom` table.

```{sql, connection=con}
SELECT * FROM ecom
```

## Limit

If you have a large table with thousands of rows, returning all the records
will take time. Use LIMIT to specify the number of records to return.

```{sql, connection=con}
SELECT * FROM ecom limit 10
```

## Distinct

A column in a table may often contain many duplicate values; and we might be interested only in the 
distinct/unique values. In such cases, we can use the SELECT DISTINCT statement to return only distinct values.

```{sql, connection=con}
SELECT distinct referrer FROM ecom
```

## Filter

Now that we know how to select columns, let us focus on filtering data. In SQL, the WHERE keyword is used to extract only those records that fulfill a specified condition. Data filter based on both text and numeric values in a table. Below are a few comparison operators we can use:

- `=` equal
- `<>` not equal
- `<` less than
- `>` greater than
- `<=` less than or equal to
- `>=` greater than or equal to


The following SQL statement filters all rows from the `ecom` table where the `duration` field
is greater than 300.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE duration > 300
```

Let us filter data based on a text value. In the following example, we filter all rows 
from the `ecom` table where the `device` used is `mobile`.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE device == 'mobile'
```

## And, Or & Not

The WHERE clause can be combined with other operators such as  

- AND - displays a record if all the conditions separated by AND is TRUE
- OR  - displays a record if any of the conditions separated by OR is TRUE
- NOT - displays a record if the condition(s) is NOT TRUE

to filter data based on more than one condition or to create more complex conditions.

In the following example, we filter all the rows from the `ecom` table where
`n_visit` (visit count) is greater than 3 and `duration` (time spent on the site)
is greater than 100. We use AND to create multiple conditions.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE n_visit > 3 AND duration > 100
```

In the next example, we will use both AND & OR. Our goal is to filter all 
rows from the `ecom` table that follow the below conditions:

- `n_visit` (visit count) is either equal to 3 or 5
- `device` used to visit the website is either `mobile` or `tablet`

```{sql, connection=con}
SELECT * 
FROM ecom WHERE (n_visit == 5 OR n_visit == 3)  
AND (device = 'mobile' OR device = 'tablet')
```

## BETWEEN

The BETWEEN operator selects values within a given range and is inclusive: begin and end values are included. The values can be numbers, text, or dates. In the following example, we filter rows from 
the `ecom` table where the visit count is between 1 and 3, and the `device` used to visit the website is
`mobile`.

```{sql, connection=con}
SELECT * 
FROM ecom
WHERE n_visit BETWEEN 1 AND 3 AND device = 'mobile'
```

## IN

The IN operator allows us to specify multiple values in a WHERE clause. It is a shorthand 
for multiple OR conditions. In the below example, we filter rows from the `ecom` table where
the visit count is either 2 or 4 or 6 or 8 or 10. Instead of using multiple OR conditions, 
we use the IN operator.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE n_visit IN (2, 4, 6, 8, 10)
```

## IS NULL

A field with a NULL value is a field with no value. If a field in a table is optional, it is possible to insert a new record or update a record without adding a value to this field. Then, the field will be saved with a NULL value. In the next example, we filter all rows from the `ecom` table where the `device` column
has NULL values.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE device IS NULL
```

## LIKE

The LIKE operator is used to search for a specific pattern in a column. There are two wildcards used in conjunction with the LIKE operator:

- % : represents zero, one, or multiple characters
- _ : represents a single character

In the following example, we filter all rows from the `ecom` table where the name of the `country` starts with `P`. We use `%` after `P` to indicate that it can be followed by any number or type of characters.

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE country LIKE 'P%'
```

Let us look at another example where we filter all rows from the `ecom` table where the name of the
country should follow the below conditions:

- name can start with any character
- the second character must be `o`
- it can have any type or number of characters after the second character

```{sql, connection=con}
SELECT * 
FROM ecom 
WHERE country LIKE '_o%'
```
