# SQL Advanced {#sql2}

## Introduction

In this chapter, we will learn to

- aggregate data
- order data
- and group data

We will use the following R packages:

- [DBI](http://readr.tidyverse.org/)
- [RSQLite](https://rstats-db.github.io/RSQLite/)
- [dbplyr](http://dbplyr.tidyverse.org/)

```{r lite1aab, echo=FALSE, eval=TRUE, results='hide', message=FALSE}
library(dbplyr)
library(DBI)
library(RSQLite)
```

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) and code can be downloaded from [here](https://gist.github.com/rsquaredacademy/691dba8e13873c648bfdb91aba65d4d7).

## Set Up

In order to be able to use the `ecom` data set for our case study, we will read 
it using the **readr** package and then copy it to an in memory database using the 
`copy_to()` function from **dplyr**.

```{r setup2, message=FALSE}
ecom <- readr::read_csv('https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv')
con <- DBI::dbConnect(RSQLite::SQLite(), ":memory:")
copy_to(con, ecom)
```

## Aggregate 

Let us combine the aggregate statements with WHERE statement to filter data.

- SUM()  : returns the total sum of a numeric column

```{sql, connection=con}
SELECT SUM(n_visit) FROM ecom
```

```{sql, connection=con}
SELECT SUM(n_visit) 
FROM ecom 
WHERE n_visit > 5
```

- AVG()  : returns the average value of a numeric column

```{sql, connection=con}
SELECT AVG(n_visit) FROM ecom
```

```{sql, connection=con}
SELECT AVG(n_visit) 
FROM ecom 
WHERE country LIKE 'P%'
```

- MAX()  : returns the largest value of the selected column

```{sql, connection=con}
SELECT MAX(n_visit) FROM ecom
```

```{sql, connection=con}
SELECT MAX(n_visit) 
FROM ecom 
WHERE device == 'tablet'
```

- MIN()  : returns the smallest value of the selected column

```{sql, connection=con}
SELECT MIN(n_visit) FROM ecom
```

```{sql, connection=con}
SELECT MIN(n_visit) 
FROM ecom 
WHERE duration BETWEEN 600 AND 900
```

## Alias 

SQL aliases are used to give a table, or a column in a table, a temporary name. They are often used to make column names more readable. An alias only exists for the duration of the query. Below are a
few examples:

```{sql, connection=con}
SELECT AVG(n_visit) AS avg_mobile 
FROM ecom 
WHERE device == 'mobile'
```

```{sql, connection=con}
SELECT MAX(n_visit) AS max_visit 
FROM ecom
```

```{sql, connection=con}
SELECT MIN(duration) AS min_duration 
FROM ecom
```

## Order By

The ORDER BY keyword is used to sort the records in ascending or descending order. By default, the records are sorted in ascending order. Use the DESC keyword if you want to sort the records in descending order, 

```{sql, connection=con}
SELECT * 
FROM ecom 
ORDER BY country
```

```{sql, connection=con}
SELECT *  
FROM ecom 
ORDER BY duration
```

```{sql, connection=con}
SELECT * 
FROM ecom 
ORDER BY n_visit DESC
```

## Group By 

The GROUP BY statement is used with aggregate functions (COUNT, MAX, MIN, SUM, AVG) to group the result by one or more columns.

```{sql, connection=con}
SELECT device, 
count(*) AS visits 
FROM ecom 
GROUP BY device 
ORDER by visits DESC
```

```{sql, connection=con}
SELECT device, MAX(duration) AS max_duration 
FROM ecom 
GROUP BY device 
ORDER by max_duration DESC
```
