# JOINs {#joins}

```{r join-libs, echo=FALSE, message=FALSE}
library(dplyr)
library(DBI)
library(RSQLite)
library(dbplyr)
knitr::opts_chunk$set(R.options=list(max.print=50))
```

## Introduction

In this chapter, we will learn to:

- combine rows from two tables with `INNER JOIN`
- keep unmatched rows with `LEFT JOIN`
- avoid the SQLite `RIGHT` / `FULL OUTER JOIN` trap
- do the same joins from R with dbplyr `inner_join()` / `left_join()`
- filter with `semi_join()` and `anti_join()`

We will use the following R packages:

- [DBI](https://dbi.r-dbi.org/)
- [RSQLite](https://rsqlite.r-dbi.org/)
- [dbplyr](http://dbplyr.tidyverse.org/)

All the data sets used in this chapter can be found [here](https://github.com/rsquaredacademy/datasets) and code can be run from `code/ch5-joins.R` in this repo.

## Set Up

So far we have worked with a single table. Real databases keep related data in
separate tables. Our `data/ecom.sqlite` file holds two small related tables
besides `ecom`:

- `customers`: 20 rows, one per customer (`customer_id` primary key, `country`)
- `orders`: 20 rows, one per order (`order_id` primary key, `customer_id`
  foreign key into `customers`, `amount`)

Six customers (4, 16--20) have placed no orders, and one order points at a
`customer_id` (99) that does not exist. These deliberate mismatches make the
difference between join types visible.

Let us connect directly to the SQLite file:

```{r join-setup, message=FALSE}
con <- DBI::dbConnect(RSQLite::SQLite(), "data/ecom.sqlite")
```

A quick look at both tables:

```{sql, connection=con}
SELECT * FROM customers LIMIT 5
```

```{sql, connection=con}
SELECT * FROM orders LIMIT 5
```

## INNER JOIN

`INNER JOIN` returns only the rows that match in **both** tables. Customers
without orders, and orders without customers, are dropped.

```{sql, connection=con}
SELECT c.customer_id, c.country, o.order_id, o.amount
FROM customers c
INNER JOIN orders o ON c.customer_id = o.customer_id
ORDER BY c.customer_id, o.order_id
LIMIT 6
```

How many of the 20 orders survive? Nineteen -- order 120 belongs to the
non-existent customer 99 and is excluded:

```{sql, connection=con}
SELECT COUNT(*) AS n FROM customers c
INNER JOIN orders o ON c.customer_id = o.customer_id
```

## LEFT JOIN

`LEFT JOIN` keeps **every** row from the left table and fills `NULL` where the
right table has no match. This is how you find customers who never ordered:

```{sql, connection=con}
SELECT c.customer_id, c.country, o.order_id, o.amount
FROM customers c
LEFT JOIN orders o ON c.customer_id = o.customer_id
WHERE o.order_id IS NULL
ORDER BY c.customer_id
```

The full left join has 25 rows: 19 matched order rows plus one `NULL` row for
each of the 6 customers without orders.

## The SQLite RIGHT / FULL Trap

Some tutorials teach `RIGHT JOIN` or `FULL OUTER JOIN`. Do not use them with
RSQLite: the SQLite versions bundled with RSQLite (including recent ones)
reject both with a syntax error such as `near "RIGHT": syntax error`.
Postgres and other production databases accept them, so queries written
against SQLite will fail silently only when you move them.

The portable workaround is to swap the tables and use `LEFT JOIN`:

- instead of `A RIGHT JOIN B`, write `B LEFT JOIN A`
- instead of `A FULL OUTER JOIN B`, union a `LEFT JOIN` in each direction

When in doubt, `LEFT JOIN` is the only outer join you need for this book.

## JOINs with dbplyr

The same joins work from R without writing SQL. Reference both tables and use
the dplyr join verbs:

```{r join-dbplyr, message=FALSE}
customers <- dplyr::tbl(con, "customers")
orders <- dplyr::tbl(con, "orders")

inner_join(customers, orders, by = "customer_id")
```

Use `show_query()` to see the SQL dbplyr generates -- it is the `INNER JOIN`
you just wrote by hand:

```{r join-show-inner, collapse=TRUE}
q_inner <- inner_join(customers, orders, by = "customer_id")
dplyr::show_query(q_inner)
```

The left join works the same way:

```{r join-show-left, collapse=TRUE}
q_left <- left_join(customers, orders, by = "customer_id")
dplyr::show_query(q_left)
```

## semi_join and anti_join

Two filtering joins have no direct SQL keyword but are everyday tools:

- `semi_join()` keeps rows of `customers` that **have** a match in `orders`
  (14 customers). The generated SQL uses `WHERE EXISTS`.
- `anti_join()` keeps rows of `customers` that **lack** a match in `orders`
  (the same 6 customers from the `LEFT JOIN` check above). The generated SQL
  uses `WHERE NOT EXISTS`.

```{r join-semi, collapse=TRUE}
q_semi <- semi_join(customers, orders, by = "customer_id")
dplyr::show_query(q_semi)
dplyr::collect(q_semi)
```

```{r join-anti, collapse=TRUE}
q_anti <- anti_join(customers, orders, by = "customer_id")
dplyr::show_query(q_anti)
dplyr::collect(q_anti)
```

Rule of thumb: use `semi_join()` / `anti_join()` when you want to filter one
table by membership in another without adding the other table's columns.

```{r join-close}
dbDisconnect(con)
```
