# DuckDB + Parquet {#sec-duckdb}

[SQLite](#sec-dbi) was the right default for learning: zero setup, same DBI
code everywhere. [DuckDB](https://duckdb.org/) is its analytical twin — an
in-process engine (no server, like SQLite) built for fast aggregate queries
over large files. The pitch for R users: the DBI code you already know runs
unchanged, and you can query Parquet or CSV files directly without loading
them into memory.

> All chunks in this appendix are display-only (`eval=FALSE`): they need
> `install.packages("duckdb")` once, which the book's automated builds skip.
> Run them locally — every line works as written. Runnable code lives in
> `code/ch7-duckdb.R` (skips gracefully when `duckdb` is not installed).

## Connect: one Line Changes

```{r duck-connect, eval=FALSE}
# install.packages("duckdb")  # once; high-speed analytical engine
con_duck <- DBI::dbConnect(duckdb::duckdb())
DBI::dbWriteTable(con_duck, "ecom", ecom)
DBI::dbGetQuery(con_duck, "SELECT referrer, AVG(duration) FROM ecom GROUP BY referrer")
```

Compare with Chapter 1: `RSQLite::SQLite()` became `duckdb::duckdb()`, and
nothing else moved. `dbListTables()`, `dbGetQuery()`, `dbWriteTable()`,
`dbDisconnect()` — the whole DBI vocabulary carries over, which is exactly
why the book taught SQLite first.

## Query Files Without Loading Them

DuckDB's signature trick: `FROM 'file'` reads Parquet or CSV straight off
disk, pushing filters and aggregates down so R never holds the full data:

```{r duck-parquet, eval=FALSE}
# Aggregate a 2 GB Parquet file from ~50 MB of R memory
DBI::dbGetQuery(con_duck, "
  SELECT country, SUM(amount) AS revenue
  FROM 'orders.parquet'
  WHERE year = 2025
  GROUP BY country
  ORDER BY revenue DESC
  LIMIT 10")

# Same for CSVs — handy for the raw logs before they become Parquet
DBI::dbGetQuery(con_duck, "
  SELECT COUNT(*) AS n FROM read_csv('logs/*.csv', header = true)")
```

Write results back out the same way (`COPY ... TO 'out.parquet'`), and you
have a full extract-transform pipeline with no database server and no RAM
anxiety. Pair with [Arrow](https://arrow.apache.org/docs/r/) (`arrow::read_parquet()`,
`arrow::write_parquet()`) when you want the files in R as well as SQL.

## dbplyr on DuckDB

Lazy evaluation works exactly as in Chapter 2 — `tbl()`, verbs,
`show_query()`, `collect()` — with DuckDB as the execution engine:

```{r duck-dbplyr, eval=FALSE}
e <- dplyr::tbl(con_duck, "ecom")
q <- e |>
  dplyr::filter(duration > 300) |>
  dplyr::group_by(device) |>
  dplyr::summarise(avg = mean(duration))
dplyr::show_query(q)  # DuckDB SQL this time, same dplyr code
dplyr::collect(q)
```

Two habits carry over from Chapter 2: use `show_query()` to check what
dbplyr generates (window functions and custom SQL still deserve hand-written
queries via `dplyr::sql()`), and prefer `dplyr::compute()` over `collect()`
for intermediate results too big for memory — `compute()` materializes them
as a DuckDB table instead of an R tibble.

```{r duck-close, eval=FALSE}
DBI::dbDisconnect(con_duck)
```

## SQLite or DuckDB?

| | SQLite (Chapters 1–5) | DuckDB (this appendix) |
|---|---|---|
| Shape | Row-oriented, transactional | Column-oriented, analytical |
| Best at | Lookups, inserts, app state | Aggregates, scans, Parquet |
| Files | `.sqlite` | `.parquet`, `.csv`, in-memory |
| Scale | Fits comfortably | Bigger than RAM, via pushdown |
| R path | `RSQLite::SQLite()` | `duckdb::duckdb()` |

Default to SQLite for application data and DuckDB for analysis. When your
analysis outgrows DuckDB on one machine, the next step up is
[R4DS Ch. 21](https://r4ds.hadley.nz/databases) — the funnel this book was
built to feed.
