# Production R Patterns {#sec-production}

So far every chapter used an in-memory SQLite database. That was deliberate:
zero setup, everything reproducible. Real jobs look different — a Postgres
server with credentials, concurrent writers, and tables far bigger than memory.

The good news: almost everything transfers. `DBI` is a common interface, so
`dbGetQuery()`, `dbWriteTable()`, and `dbDisconnect()` work the same everywhere.
This appendix covers the six things that change when you leave SQLite.

We will use the following R packages:

- [DBI](https://dbi.r-dbi.org/)
- [RSQLite](https://rsqlite.r-dbi.org/)

Sections that need a live server are display-only (`eval=FALSE`) — read them,
do not run them. Sections that run on SQLite execute normally.

## Connection Strings

Only the `dbConnect()` call changes per backend. Everything after it is
identical DBI code:

```{r pg-connect, eval=FALSE}
# Postgres (production default for most R teams)
con <- DBI::dbConnect(RPostgres::RPostgres(),
  host = "db.internal", dbname = "analytics",
  user = Sys.getenv("DB_USER"), password = Sys.getenv("DB_PASS"))

# MariaDB / MySQL
con <- DBI::dbConnect(RMariaDB::MariaDB(),
  host = "db.internal", dbname = "analytics",
  user = Sys.getenv("DB_USER"), password = Sys.getenv("DB_PASS"))

# Anything with an ODBC driver (SQL Server, Oracle, Snowflake)
con <- DBI::dbConnect(odbc::odbc(), dsn = "warehouse",
  uid = Sys.getenv("DB_USER"), pwd = Sys.getenv("DB_PASS"))
```

For the full backend matrix (BigQuery, Snowflake, Oracle DSNs, enterprise
security), see [Posit Databases using R](https://solutions.posit.co/connections/db/).
This book stays backend-agnostic on purpose: learn the patterns here once,
then look up one connection string.

## Credentials: env vars + keyring

Never paste passwords into scripts. Two rules, in order of preference:

```{r creds, eval=FALSE}
# 1. Environment variables (works everywhere: laptop, CI, Posit Connect)
Sys.setenv(DB_USER = "analyst")  # once per session, never committed
con <- DBI::dbConnect(RPostgres::RPostgres(), password = Sys.getenv("DB_PASS"))

# 2. OS credential store for interactive work (macOS Keychain, Windows
#    Credential Manager, Linux Secret Service)
keyring::key_set("warehouse", "analyst")  # prompts once, stores securely
pwd <- keyring::key_get("warehouse", "analyst")
```

If a password appears in a committed file, rotate it — editing history does
not un-leak it.

## Parameterized Queries

Build SQL with placeholders, not `paste0()`. String-pasting user input into
queries is how SQL injection happens (especially in Shiny apps). This section
executes — placeholders work on SQLite too:

```{r param-setup, echo=FALSE, message=FALSE}
library(DBI)
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(), ":memory:")
web_path <- if (file.exists("data/web.csv")) "data/web.csv" else "https://raw.githubusercontent.com/rsquaredacademy/datasets/master/web.csv"
ecom <- readr::read_csv(web_path, show_col_types = FALSE)
dplyr::copy_to(con, ecom)
```

```{r sqlinterp}
device <- "mobile"
q <- DBI::sqlInterpolate(con,
  "SELECT COUNT(*) AS n FROM ecom WHERE device = ?dev",
  dev = device)
q
dbGetQuery(con, q)
```

`sqlInterpolate()` quotes and escapes `?dev` for the active backend, so the
same code is safe on Postgres. For composing queries programmatically, the
equivalent is `glue::glue_sql("... {dev}", .con = con)` — note `.con`, which
tells `glue` which dialect to quote for.

## Transactions

When several writes must succeed or fail together (debit one account, credit
another), wrap them in a transaction. `dbWithTransaction()` commits on success
and rolls back automatically if any step errors:

```{r txn}
dbWriteTable(con, "trial", data.frame(id = 1:3, x = c("a", "b", "c")))
dbWithTransaction(con, {
  dbExecute(con, "INSERT INTO trial (id, x) VALUES (4, 'd')")
  dbExecute(con, "DELETE FROM trial WHERE id = 1")
})
dbGetQuery(con, "SELECT * FROM trial ORDER BY id")
```

A failed transaction leaves the table untouched — verify by forcing an error
inside the block and re-reading the table. Prefer `dbWithTransaction()` over
manual `dbBegin()` / `dbCommit()` / `dbRollback()`: the manual trio leaks open
transactions when an error interrupts your script.

## Writing Data at Scale

`copy_to()` (used throughout this book) is a demo convenience: it uploads an
in-memory data frame with guessed column types. For real loads:

```{r writes, eval=FALSE}
# Declare types explicitly instead of letting the database guess
DBI::dbCreateTable(con, "events",
  fields = c(id = "INTEGER", ts = "TIMESTAMP", label = "TEXT"))

# Control large uploads: overwrite once, append after
DBI::dbWriteTable(con, "events", january, overwrite = TRUE)
DBI::dbWriteTable(con, "events", february, append = TRUE)

# Speed repeated lookups with an index (or a view for saved queries)
DBI::dbExecute(con, "CREATE INDEX idx_events_ts ON events (ts)")
```

Rule of thumb: `copy_to()` for classroom demos, `dbWriteTable()` with
`overwrite` / `append` for pipelines, `dbCreateTable()` + explicit types when
the table must survive longer than your session.

## Connections in Shiny: pool

A Shiny app that calls `dbConnect()` per session exhausts the database under
load and leaks idle connections. The [pool](https://rstudio.github.io/pool/)
package checks connections out per reactive flush and returns them after:

```{r pool, eval=FALSE}
pool <- pool::dbPool(RPostgres::RPostgres(), dbname = "analytics",
  host = "db.internal", user = Sys.getenv("DB_USER"))
# use pool exactly like con, then on app stop:
pool::poolClose(pool)
```

If you take one thing from this appendix into production Shiny work, take
`pool`.

```{r prod-close}
dbDisconnect(con)
```

Further reading: `vignette("DBI-advanced")` (transactions, quoting, bulk
operations) and the [DBI specification](https://dbi.r-dbi.org/) for backend
authors. Runnable code for this appendix lives in `code/ch6-production.R`
(server sections included as comments — uncomment with your credentials).
